A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-09-03 main.tex
2020-09-03 meeting edits

This mechanism of environmental confounding is summarized in the directed acyclic graph displayed in Figure \ref{fig:ps_env}. The undirected dashed line between population stratification and environmental exposure indicates that while we assume these elements are correlated, we do not assume any causal direction. Otherwise stated, we do not assume that environmental factors directly influence population-specific allele frequencies, or vice versa, only that the two may be related. In the context of our air quality example, this means that we assume air quality is not directly altering individuals' allele frequencies, nor are allele frequencies causing individuals to move to regions of better or worse air quality. Genetic relatedness and environmental effects may become associated over time for any number of complex reasons beyond the scope of this paper. However, as we will discuss, many methods of correcting for population structure rely on this relationship and do not formally differentiate between effects of population structure and those of environment. Such methods assume that genetic relatedness serves as an accurate proxy for environmental exposures, which may or may not be the case depending on the exposure and population in question. \pb{is this an important point?} \anna{hmm... maybe? robust to this assumption: if it's not a good proxy it's not going to cause confounding}

@@

Using the result of the eigendecomposition \pb{in what sense is this an eigendecomposition?  I would describe it as a Cholesky decomposition} from \eqref{eqn:k}, $\bK = p^{-1} \bR \bR^\T$ , the LMM model \eqref{eqn:lmm_reg} can be written equivalently as

@@

@@

In carrying out simulation studies of the LMM regression model \eqref{eqn:lmm_reg}, a common practice is to draw a new value of the quantity $\bX\bgamma$ from a $N(\boldsymbol{0}, \sigmagg \bK)$ for each replication of the analysis \citep{Rakitsch2012, bhatnagar2019simultaneous}.  We argue that this is somewhat misleading from the perspective of understanding the statistical properties of the method. Averaging over situations in which the environmental confounding is constantly changing direction and magnitude has the effect of ``washing out'' bias in $\bbetaHat$. To illustrate this concept, consider the low-dimensional ordinary least squares (OLS) estimates of $\bbeta$ based on model \eqref{eqn:lmm_reg} where, without loss of generatlity, we assume $\bX_0 \bbeta_0 = \boldsymbol{0}$ for ease of notation: 
\begin{align}
    \mathbb{E}(\bbetaHat) &= (\bX^\T \bX)^{-1} \bX^\T \mathbb{E}(\by) \notag \\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{E}(\bX\bbeta + \bX\bgamma + \beps) \label{eqn:expectation}\\
    &\quad \notag\\
    \mathbb{V}(\bbetaHat) &= (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\by) \notag\\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\bX\bbeta + \bX\bgamma + \beps) \notag\\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\bX\bgamma + \beps) \bX  (\bX^\T \bX)^{-1} \label{eqn:variance}.
\end{align}
Under the assumption that environmental effects arise from a random process, $\bgamma \sim N(\boldsymbol{0}, p^{-1} \sigmagg \bI)$, equations \eqref{eqn:expectation} and \eqref{eqn:variance} yield $\mathbb{E}(\bbetaHat) = \bbeta$, and $\mathbb{V}(\bbetaHat) = p^{-1}\sigmagg\bI + \sigmaee (\bX^\T \bX)^{-1}$, respectively. Consider our earlier example involving confounding due to air quality \pb{I feel that this should be moved up to before the mathematical derivations}. Simulating $\bgamma$ as a random quantity for each replication corresponds to subpopulation A being exposed to good air quality in one replication, and bad air quality in the next. Reporting the average across these situations is inconsistent with the question we are truly interested in, namely: what would happen if we took repeated samples from populations A and B, when these populations are subject to different environmental exposures? Those repeated samples would show a systematic bias due to the difference in air quality, but this is masked if $\bgamma$ is assumed to be random. Treating $\bgamma$ as a fixed quantity so that environmental effects are consistent across replications more closely approximates environmental confounding. Under the assumption that the quantity $\bgamma$ is fixed, equations \eqref{eqn:expectation} and \eqref{eqn:variance} yield $\mathbb{E}(\bbetaHat) = \bbeta + \bgamma, \mathbb{V}(\bbetaHat) = \sigmaee (\bX^\T \bX)^{-1}$, allowing us to quantify estimation bias due to environmental heterogeneity.

Although we derive these quantities under model \eqref{eqn:lmm_reg}, where $\bZ = \bX$, these results also hold for the more general model \eqref{eqn:model}, where $\bZ \ne \bX$. We see this by decomposing $\bZ \bgamma$ into components in and orthogonal to the column space of the projection matrix of $\bX$.  Let $\boldsymbol{P}_{\bX} = \bX (\bX^\T \bX)^{-1} \bX^\T$ denote the projection matrix of $\bX$, with $\mathcal{C}(\boldsymbol{P}_{\bX})$ and $\mathcal{N}(\boldsymbol{P}_{\bX})$ denoting the column and null spaces of $\boldsymbol{P}_{\bX}$, respectively. Then there exist $\boldsymbol{\tau}, \boldsymbol{\psi}: \bZ \bgamma = \bX \boldsymbol{\tau + \psi}$, where $\bX \boldsymbol{\tau} \in \mathcal{C}(\boldsymbol{P}_{\bX})$ and $\boldsymbol{\psi} \in \mathcal{N}(\boldsymbol{P}_{\bX})$.  This in turn means that
\begin{align*}
    \by &= \bX\bbeta + \bZ\bgamma + \beps\\
    &= \bX\bbeta + \bX \boldsymbol{\tau} + \boldsymbol{\psi} + \beps\\
    &=  \bX \bbeta{^*} + \beps{^*};
\end{align*}
note that $\boldsymbol{\tau}$ quantifies bias in the estimated $\bbetaHat^*$ and $\boldsymbol{\psi}$ is absorbed by the residual error term. In other words, any model specified in terms of $\bZ$ and $\bgamma$ is equivalent to a parameterization using $\bX$ and $\boldsymbol{\tau}$.

>>>>

This mechanism of environmental confounding is summarized in the directed acyclic graph displayed in Figure \ref{fig:ps_env}. The undirected dashed line between population stratification and environmental exposure indicates that while we assume these elements are correlated, we do not assume any causal direction. Otherwise stated, we do not assume that environmental factors directly influence population-specific allele frequencies, or vice versa, only that the two may be related. In the context of our air quality example, this means that we assume air quality is not directly altering individuals' allele frequencies, nor are allele frequencies causing individuals to move to regions of better or worse air quality. Genetic relatedness and environmental effects may become associated over time for any number of complex reasons beyond the scope of this paper. However, as we will discuss, many methods of correcting for population structure rely on this relationship and do not formally differentiate between effects of population structure and those of environment. Such methods assume that genetic relatedness serves as an accurate proxy for environmental exposures, which may or may not be the case depending on the exposure and population in question.

@@

Using the result of the decomposition from \eqref{eqn:k}, $\bK = p^{-1} \bR \bR^\T$ , the LMM model \eqref{eqn:lmm_reg} can be written equivalently as

@@

\anna{(1) fixed vs. random, (2) decomposition}

@@

\anna{qualitative explanation, then quantitative}

In carrying out simulation studies of the LMM regression model \eqref{eqn:lmm_reg}, a common practice is to draw a new value of the quantity $\bX\bgamma$ from a $N(\boldsymbol{0}, \sigmagg \bK)$ for each replication of the analysis \citep{Rakitsch2012, bhatnagar2019simultaneous}.  We argue that this is somewhat misleading from the perspective of understanding the statistical properties of the method. Averaging over situations in which the environmental confounding is constantly changing direction and magnitude has the effect of ``washing out'' bias in $\bbetaHat$. To illustrate this concept, consider the low-dimensional ordinary least squares (OLS) estimates of $\bbeta$ based on model \eqref{eqn:lmm_reg} where, without loss of generatlity, we assume $\bX_0 \bbeta_0 = \boldsymbol{0}$ for ease of notation: 
\begin{align}
    \mathbb{E}(\bbetaHat) &= (\bX^\T \bX)^{-1} \bX^\T \mathbb{E}(\by) \notag \\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{E}(\bX\bbeta + \bX\bgamma + \beps) \label{eqn:expectation}\\
    &\quad \notag\\
    \mathbb{V}(\bbetaHat) &= (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\by) \notag\\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\bX\bbeta + \bX\bgamma + \beps) \notag\\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\bX\bgamma + \beps) \bX  (\bX^\T \bX)^{-1} \label{eqn:variance}.
\end{align}
Under the assumption that environmental effects arise from a random process, $\bgamma \sim N(\boldsymbol{0}, p^{-1} \sigmagg \bI)$, equations \eqref{eqn:expectation} and \eqref{eqn:variance} yield $\mathbb{E}(\bbetaHat) = \bbeta$, and $\mathbb{V}(\bbetaHat) = p^{-1}\sigmagg\bI + \sigmaee (\bX^\T \bX)^{-1}$, respectively. Consider our earlier example involving confounding due to air quality \pb{I feel that this should be moved up to before the mathematical derivations}. Simulating $\bgamma$ as a random quantity for each replication corresponds to subpopulation A being exposed to good air quality in one replication, and bad air quality in the next. Reporting the average across these situations is inconsistent with the question we are truly interested in, namely: what would happen if we took repeated samples from populations A and B, when these populations are subject to different environmental exposures? Those repeated samples would show a systematic bias due to the difference in air quality, but this is masked if $\bgamma$ is assumed to be random. Treating $\bgamma$ as a fixed quantity so that environmental effects are consistent across replications more closely approximates environmental confounding. Under the assumption that the quantity $\bgamma$ is fixed, equations \eqref{eqn:expectation} and \eqref{eqn:variance} yield $\mathbb{E}(\bbetaHat) = \bbeta + \bgamma, \mathbb{V}(\bbetaHat) = \sigmaee (\bX^\T \bX)^{-1}$, allowing us to quantify estimation bias due to environmental heterogeneity. 
\anna{talk about a low-rank approximation of xtx-inv for a unique solution}
Although we derive these quantities under model \eqref{eqn:lmm_reg}, where $\bZ = \bX$, these results also hold for the more general model \eqref{eqn:model}, where $\bZ \ne \bX$. We see this by decomposing $\bZ \bgamma$ into components in and orthogonal to the column space of the projection matrix of $\bX$.  Let $\boldsymbol{P}_{\bX} = \bX (\bX^\T \bX)^{-1} \bX^\T$ denote the projection matrix of $\bX$, with $\mathcal{C}(\boldsymbol{P}_{\bX})$ and $\mathcal{N}(\boldsymbol{P}_{\bX})$ denoting the column and null spaces of $\boldsymbol{P}_{\bX}$, respectively. Then there exist $\boldsymbol{\tau}, \boldsymbol{\psi}: \bZ \bgamma = \bX \boldsymbol{\tau + \psi}$, where $\bX \boldsymbol{\tau} \in \mathcal{C}(\boldsymbol{P}_{\bX})$ and $\boldsymbol{\psi} \in \mathcal{N}(\boldsymbol{P}_{\bX})$.  This in turn means that
\begin{align*}
    \by &= \bX\bbeta + \bZ\bgamma + \beps\\
    &= \bX\bbeta + \bX \boldsymbol{\tau} + \boldsymbol{\psi} + \beps\\
    &=  \bX \bbeta{^*} + \beps{^*};
\end{align*}
note that $\boldsymbol{\tau}$ quantifies bias in the estimated $\bbetaHat^*$ and $\boldsymbol{\psi}$ is absorbed by the residual error term. In other words, any model specified in terms of $\bZ$ and $\bgamma$ is equivalent to a parameterization using $\bX$ and $\boldsymbol{\tau}$.
